---
title: "Covid91 vaccine study Final2024"
author: "Ashley Saunders"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
OldData <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- OldData %>% mutate(infected = recode(infected, covid19 = "covid", none = "OK" )) %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
FinalData$treatment <- relevel(FinalData$treatment,ref="placebo")
#FinalData$infected <- relevel(FinalData$infected,ref="OK")
Male = filter(FinalData, sex =="M")
Female = filter(FinalData, sex =="F")
Gay = filter(FinalData, LGBTQ =="gay")
DrugUser = filter(FinalData, DrugUser =="yes")
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

## Overall approach

1. **Descriptive Analysis**: Summarize data perceptibly then numerically.
2. **Inferential Analysis**: Perform hypothesis testing:
   - \(H_0\): The vaccine has no impact on infection rates.
   - \(H_A\): The vaccine reduces infection rates.
3. **Effectiveness Calculation**:
   \[
   \text{Vaccine Effectiveness (VE)} = 1 - \frac{\text{Risk}_{\text{vaccine}}}{\text{Risk}_{\text{placebo}}}
   \]
4. **Effectivness Confidence Interval**: low to high

## Males

### Summary of Males
```{r}
male_summary <- Male %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)
male_summary <- male_summary %>%
  mutate(total = rowSums(select(., -treatment)))
male_column_totals <- male_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")
male_summary <- bind_rows(male_summary, male_column_totals)
male_summary
```

For men, this is the information that has been provided at this point. 18392 men are present. 18037 are acceptable but 355 have COVID. Let's add a barplot for a visual depiction since this is insufficient to demonstrate a trend.

### Visual Results
```{r}
# Bar plot

# Summarize the data for Males by infection status and treatment
male_summary_data <- Male %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
male_plot_count <- ggplot(male_summary_data, aes(x = infected, y = count, fill = treatment))
male_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Males Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Males") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
male_plot_percent <- ggplot(male_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
male_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Males Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Males") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The number of males infected with COVID-19 is displayed in the barplots above. On both graphs, the findings are comparable to the placebo, but visually, males who receive the immunization have lower COVID than the placebo. They still differ visually, but in order to determine whether the difference is significant, we must perform a statistical study.

### Numerical Results
```{r}
table1 <- xtabs(~infected + treatment, data=Male)
rowPerc(table1)
colPerc(table1)
```

According to the numerical data, 73.8% of the diseased males had received the placebo, whereas 26.2% had received the vaccine. 49.99% had received the vaccine, and 50.01% had received the placebo for those who are acceptable. Overall, compared to 1.02% of males receiving the vaccine, 2.82% of guys receiving the placebo had COVID, suggesting that the immunization lowers infection rates. These results imply that the immunization is successful in reducing COVID-19 in males. This allows us to correlate it to the barplots that demonstrate its success.

### Inferential Results
```{r}
chisq.test(table1)
chisqtestGC(table1)
Malefish <- fisher.test(table1)
Malefish
```

Fisher's Exact Test yields a very small p-value, 0.00000000000000022, which is the main result to pay attention to. This indicates that it is highly improbable that the observed variations in infection rates between the vaccine and placebo groups happened by accident. We reject the null hypothesis, which holds that infection rates are unaffected by therapy. Males who received the vaccine had an approximately 2.8-fold higher risk of contracting COVID-19 than those who received a placebo, as indicated by the odds ratio of roughly 2.82. These findings suggest that the immunization has little effect on preventing COVID-19 in men.

```{r}
# Calculate effectiveness
male_effect <- Malefish$p.value / (Malefish$p.value + 1)
effectiveness_male <- (1 - 1 / Malefish$estimate) * 100
print(effectiveness_male)

# Calculate effectiveness confidence interval
male_low <- (1 - 1 / Malefish$conf.int[1]) * 100
male_hi <- (1 - 1 / Malefish$conf.int[2]) * 100
options(digits = 2)

#print 
male_low
male_hi
```

### `r effectiveness_male`% Effectiveness with Confidence Interval: `r male_low`%  to `r male_hi`%

The COVID-19 vaccine for men is effective, with an efficacy of 64.48%, which is above the 50% threshold for success. The confidence interval ranges from 54.75% to 72.32%, further supporting its effectiveness in reducing infection rates among males. The null hypothesis (no effect) is rejected, and the alternative hypothesis (vaccine lowers infection rates) is accepted, based on the low p-value from Fisher's Exact Test and an odds ratio of 2.82. This means vaccinated men are nearly 2.8 times more likely to avoid COVID-19 compared to those who received a placebo. The confidence interval further confirms the vaccine’s strong protective effect.

## Females

### Summary of Females
```{r}
female_summary <- Female %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)
female_summary <- female_summary %>%
  mutate(total = rowSums(select(., -treatment)))
female_column_totals <- female_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")
female_summary <- bind_rows(female_summary, female_column_totals)
female_summary
```

For the women, here is the information that has been provided thus far. The number of females is 18328. While 17963 are okay, 365 have COVID. Since this is insufficient to demonstrate a trend, let's now include a barplot for a visual aid.

### Visual Results
```{r}
# Bar plot

# Summarize the data for Females by infection status and treatment
female_summary_data <- Female %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
female_plot_count <- ggplot(female_summary_data, aes(x = infected, y = count, fill = treatment))
female_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Females Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Females") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
female_plot_percent <- ggplot(female_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
female_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Females Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Females") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The number of females infected with COVID-19 is displayed in the barplots above. On both graphs, the results are comparable to the placebo, but visually, females who receive the immunization have lower COVID than the placebo. They still differ visually, but in order to determine whether the difference is significant, we must perform a statistical study.

### Numerical Results
```{r}
table2 <- xtabs(~infected + treatment, data=Female)
rowPerc(table2)
colPerc(table2)
```

According to the numerical data, 35% of the diseased females had received the vaccine, whereas 65% had received the placebo. Half of individuals who are doing well had received the vaccine, and the other half had received the placebo. Overall, 1.4% of females receiving the immunization had COVID, but 2.6% of those receiving the placebo did, suggesting that the vaccine lowers infection rates. These results imply that the immunization is successful in reducing female COVID-19 rates. This is consistent with the barplots demonstrating the efficacy of the vaccine.

### Inferential Results
```{r}
chisq.test(table2)
chisqtestGC(table2)
Femalefish <- fisher.test(table2)
Femalefish
```

The key finding is from Fisher's Exact Test, which shows that the difference in infection rates between the vaccine and placebo groups is extremely unlikely to be due to chance, with a very small p-value of 0.00000001. This leads us to reject the null hypothesis, which suggests that the therapy has no effect on infection rates. The odds ratio of about 1.9 means that women who received the vaccine had nearly twice the risk of getting COVID-19 compared to those who got the placebo. Therefore, the vaccine seems ineffective at preventing COVID-19 in females.

```{r}
# Calculate effectiveness
female_effect <- Femalefish$p.value / (Femalefish$p.value + 1)
effectiveness_female <- (1 - 1 / Femalefish$estimate) * 100
print(effectiveness_female)

# Calculate effectiveness confidence interval
female_low <- (1 - 1 / Femalefish$conf.int[1]) * 100
female_hi <- (1 - 1 / Femalefish$conf.int[2]) * 100
options(digits = 2)

#print 
female_low
female_hi
```

### `r effectiveness_female`% Effectiveness with Confidence Interval: `r female_low`%  to `r female_hi`%

The vaccine's efficacy for females is shown above. A vaccine is considered effective if its efficacy is above 50%. Since the calculated efficacy is 46.13%, with a confidence interval of 32.79% to 56.98%, we conclude that the vaccine is ineffective in reducing infection rates among females. The alternative hypothesis suggested the vaccine lowers infection rates, while the null hypothesis stated it has no effect. Based on the very low p-value from Fisher's Exact Test and an odds ratio of 1.9, we reject the null hypothesis and accept the alternative. This means the vaccine actually increases the risk of infection in females. The odds ratio shows that women who got the vaccine were nearly 1.9 times more likely to contract COVID-19 than those who received the placebo. The confidence interval supports this, indicating a modest, and likely negative, impact on vaccine effectiveness.

## LGBTQ

### Summary of LGBTQ
```{r}
gay_summary <- Gay %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)
gay_summary <- gay_summary %>%
  mutate(total = rowSums(select(., -treatment)))
gay_column_totals <- gay_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")
gay_summary <- bind_rows(gay_summary, gay_column_totals)
gay_summary
```

This is the LGBTQ information that has been provided thus far. 3060 LGBTQ people are present. While 3000 are OK, 60 have COVID. Since this is insufficient to demonstrate a trend, let's now include a barplot for a visual aid.

### Visual Results
```{r}
# Bar plot

# Summarize the data for Gay by infection status and treatment
gay_summary_data <- Gay %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
gay_plot_count <- ggplot(gay_summary_data, aes(x = infected, y = count, fill = treatment))
gay_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Gay Individuals Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Individuals") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
gay_plot_percent <- ggplot(gay_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
gay_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Gay Individuals Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Individuals") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The number of LGBTQ people infected with COVID-19 is displayed in the barplots above. Compared to placebo, visually gay people who receive the immunization had lower COVID rates. Although they differ greatly visually, we must perform a statistical analysis to see whether the difference is significant.

### Numerical Results
```{r}
table3 <- xtabs(~infected + treatment, data=Gay)
rowPerc(table3)
colPerc(table3)
```

According to the numerical data, 20% of the afflicted gays had received the vaccine, whereas 80% had received a placebo. Of those who are okay, 52% had received the vaccine and 48% had received a placebo. Overall, 0.77% of gay people receiving the immunization have COVID-19, compared to 3.2% of those receiving the placebo, suggesting that the vaccine lowers infection rates. These results imply that the immunization is successful in reducing the prevalence of COVID-19 among LGBT people. This is consistent with the barplots demonstrating the efficacy of the vaccine.

### Inferential Results
```{r}
chisq.test(table3)
chisqtestGC(table3)
Gayfish <- fisher.test(table3)
Gayfish
```

The Fisher's Exact Test yields the most important finding, with a p-value of 0.0000007, which is incredibly small and suggests that the observed differences in infection rates between the vaccine and placebo groups are unlikely to have happened by accident. We reject the null hypothesis, which holds that infection rates are unaffected by therapy. Gay people who received the vaccine had an approximately 4.3-fold higher risk of contracting COVID-19 than those who received a placebo, according to the odds ratio of roughly 4.3. These findings imply that the immunization does not protect gay people from COVID-19.

```{r}
# Calculate effectiveness
gay_effect <- Gayfish$p.value / (Gayfish$p.value + 1)
effectiveness_gay <- (1 - 1 / Gayfish$estimate) * 100
print(effectiveness_gay)

# Calculate effectiveness confidence interval
gay_low <- (1 - 1 / Gayfish$conf.int[1]) * 100
gay_hi <- (1 - 1 / Gayfish$conf.int[2]) * 100
options(digits = 2)

#print 
gay_low
gay_hi
```

### `r effectiveness_gay`% Effectiveness with Confidence Interval: `r gay_low`%  to `r gay_hi`%

The efficacy of the COVID-19 vaccine for gay people is shown above. A vaccine is considered effective if its efficacy is above 50%. With an efficacy of 76.6% and a confidence interval between 55.06% and 88.74%, we can conclude that the vaccine is effective in reducing infection rates among gay people. The alternative hypothesis suggests the vaccine lowers infection rates, while the null hypothesis suggests it has no effect. The very low p-value from Fisher's Exact Test and an odds ratio of 4.3 lead us to reject the null hypothesis and accept the alternative. This means the vaccine is effective in lowering infection rates. However, the odds ratio indicates that gay people who received the vaccine had a nearly 4.3-fold higher risk of contracting COVID-19 compared to those who got the placebo. The confidence interval further supports this, confirming the vaccine has a significant protective effect for the LGBTQ community.

## Druggies

### Summary of DrugUser
```{r}
druguser_summary <- DrugUser %>%
  group_by(treatment, infected) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = infected, values_from = count, values_fill = 0)

# Add total column
druguser_summary <- druguser_summary %>%
  mutate(total = rowSums(select(., -treatment)))

# Add total row for each treatment
druguser_column_totals <- druguser_summary %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(treatment = "Total")

# Combine the totals with the main data
druguser_summary <- bind_rows(druguser_summary, druguser_column_totals)

druguser_summary
```

This is the information that drug users have been given thus far. 3060 people use drugs. While 3000 are OK, 60 have COVID. Since this is insufficient to demonstrate a trend, let's now include a barplot for a visual aid. 

### Visual Results
```{r}
# Summarize the data for DrugUser by infection status and treatment
druguser_summary_data <- DrugUser %>%
  group_by(infected, treatment) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(prcnt = count / sum(count))

# Bar plot for count of infected vs treatment
druguser_plot_count <- ggplot(druguser_summary_data, aes(x = infected, y = count, fill = treatment))
druguser_plot_count + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Drug Users Infected by Treatment",
       x = "Infection Status", y = "Count of Infected Drug Users") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()

# Bar plot for percentage of infected vs treatment
druguser_plot_percent <- ggplot(druguser_summary_data, aes(x = infected, y = prcnt * 100, fill = treatment))
druguser_plot_percent + 
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Drug Users Infected by Treatment",
       x = "Infection Status", y = "Percentage of Infected Drug Users") +
  scale_fill_manual(values = c("placebo" = "gray", "vaccine" = "blue")) +
  theme_minimal()
```

The number of drug users with COVID-19 is displayed in the barplots above. In terms of visuals, drug users who receive the immunization have higher COVID than those who receive a placebo. Even though the difference seems substantial on the surface, a statistical analysis is necessary to determine whether it is statistically significant.

### Numerical Results
```{r}
table4 <- xtabs(~infected + treatment, data=DrugUser)
rowPerc(table4)
colPerc(table4)
```

According to the numerical figures, 91.7% of infected drug users had received the vaccine, whereas 8.3% had received a placebo. 49.0% of those who are okay had received the vaccine, whereas 51.0% had received the placebo. Overall, 3.6% of drug users who received the vaccine and 0.33% of those who received the placebo got COVID, suggesting that the vaccine was linked to a greater infection risk in this population. These results imply that the vaccine might not be successful in reducing drug users' incidence of COVID-19. This is consistent with the barplots, which graphically show the increased infection rates among drug users who have received vaccinations.

### Inferential Results
```{r}
chisq.test(table4)
chisqtestGC(table4)
DrugUserfish <- fisher.test(table4)
DrugUserfish
```

The Fisher's Exact Test provides the most important finding, with a p-value of 0.000000000003, which is incredibly small and suggests that the observed differences in infection rates between the vaccine and placebo groups are highly unlikely to have happened by accident. The null hypothesis is disproved since it presupposes that infection rates are unaffected by therapy. Drug users who received the vaccine had an 11.5-fold lower risk of contracting COVID-21 than those who received a placebo, according to the odds ratio, which is roughly 0.087. According to these findings, the vaccine is quite successful in preventing COVID-19 in drug users. This is consistent with the statistical findings that indicate lower infection rates among those who have received vaccinations in this cohort.

```{r}
# Calculate effectiveness
DrugUserfish <- fisher.test(table4)
effectiveness_drug <- (1 - DrugUserfish$estimate) * 100
print(effectiveness_drug)

# Calculate effectiveness confidence level
drug_low <- (1 - DrugUserfish$conf.int[2]) * 100  
drug_hi <- (1 - DrugUserfish$conf.int[1]) * 100  
options(digits = 2)

#print
drug_low
drug_hi
```

### `r effectiveness_drug`% Effectiveness with Confidence Interval: `r drug_low`%  to `r drug_hi`%

The efficacy of the COVID-19 vaccine for drug users is shown above. A vaccine is considered effective if its efficacy is over 50%. With an efficacy of 91.3% and a confidence interval of 78.0% to 97.3%, we can conclude that the vaccine is highly effective at reducing infection rates among drug users. The alternative hypothesis suggests the vaccine lowers infection rates, while the null hypothesis says it has no effect. The very low p-value from Fisher's Exact Test and an odds ratio of 0.087 lead us to reject the null hypothesis. This means the vaccine significantly lowers infection rates for drug users. The odds ratio of 0.087 shows that vaccinated drug users have an 11.5 times lower risk of contracting COVID-19 compared to those who received the placebo. Since the odds ratio is less than 1, the vaccine is highly effective. The confidence interval further supports this, confirming the vaccine’s strong protective effect for drug users.

# Overall Results and Conclusions 

## Conclusion for Males
For males, the efficacy of the COVID vaccine is 64.48%, with a confidence interval between 54.75% and 72.32%. This conclusion is well supported by statistical tests: The null hypothesis of no vaccine impact was rejected by Fisher's Exact Test, which revealed a highly significant p-value (\(p = 2.2 \times 10^{-16}\)).Male recipients of the vaccination had a nearly 2.8-fold lower risk of infection than those receiving a placebo, according to the odds ratio of 2.82. Overall, the vaccination greatly lowers the risk of illness in males, demonstrating its efficacy in this population. 

## Conclusion for Females
With a confidence interval spanning from 32.79% to 56.98%, the effectiveness of the COVID vaccine for females is 46.13%. Additional information is provided via statistical tests: With a significant p-value (\(p = 1 \times 10^{-8}\)) from Fisher's Exact Test, the null hypothesis that there is no vaccine effect was rejected. With an odds ratio of 1.9, female recipients of the vaccine had a nearly 1.9-fold increased risk of contracting the infection in comparison to those who received a placebo. Overall, the vaccination has a limited level of efficiency and should be carefully examined for females, even though it does show some protective impact.

## Conclusion for LGBTQ
With a confidence interval spanning from 55.06% to 88.74%, the effectiveness of the COVID vaccine for gay people is 76.6%. This conclusion is well supported by statistical tests: The null hypothesis that the vaccine had no effect was rejected by the highly significant p-value (\(p = 7 \times 10^{-7}\)) obtained from Fisher's Exact Test. Gay people who received the vaccine had an almost 4.3-fold higher risk of contracting the virus than those who received a placebo, according to the odds ratio of 4.3. All things considered, the vaccination considerably lowers the risk of infection for LGBT people, demonstrating its efficacy in this demographic.

## Conclusion for Druggies
The effectiveness of the COVID vaccine for drug users is 91.3%, with a confidence interval between 78.0% and 97.3%. This conclusion is well supported by statistical tests: The null hypothesis of no vaccine impact was rejected by Fisher's Exact Test, which revealed an exceptionally significant p-value (\(p = 3 \times 10^{-12}\)). With an odds ratio of 0.087, drug users who got the vaccine had an approximate 11.5-fold lower risk of contracting the infection than those who took a placebo. Overall, the vaccine's substantial protective impact against drug users is supported by its maximal effectiveness in lowering their rates of COVID infection.



